---
title: "Supplementary Data: RT-qPCR Gene Expression and Fruit Morphometrics Data Analysis"
subtitle: "Ripening-Associated Gene Expression in Climacteric and Non-Climacteric Melon Cultivars"
author: ""
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    fig-pos: "H"
    keep-tex: false
    geometry: margin=2.5cm
    fontsize: 11pt
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

{{< pagebreak >}}

# Overview

This document provides full reproducibility documentation for the
RT-qPCR gene expression and fruit morphological data analysis
accompanying the manuscript. All raw data, scripts, and intermediate
outputs are available in the project repository.

The RT-qPCR analysis quantifies the relative expression of four
ripening-associated genes — *CmACS*, *CmACO*, *CmATH*, and *CmEREBP* —
across three developmental stages (15 DAA, 30 DAA, and maturity) in four
melon (*Cucumis melo*) cultivars (GMP, Kinaya, Melona, Tacapa Gold).
Expression is normalized to the reference gene *CmCYP* using the ΔΔCq
method (Livak & Schmittgen, 2001). The fruit morphology data are
gathered to provide supporting data in regard of melon fruit ripening
pattern.

## Software and Package Versions

```{r}
#| label: session-info
R.version.string
packageVersion("tidyverse")
packageVersion("multcompView")
packageVersion("patchwork")
```

------------------------------------------------------------------------

# RT-qPCR Data Processing

## Data Preparation

Raw Cq values and RFU amplification curves are loaded from `data/1-cq/`
and `data/2-rfu/` respectively, combined across all four cultivars, and
saved as analysis-ready `.csv` files.

The sample identifier follows the format
`[CultivarPrefix].[Stage].[BiologicalReplicate]` (e.g., `G.15.A` = GMP
cultivar, 15 DAA, biological replicate A). NTC (no-template control)
wells are retained through this step for downstream quality checking.

```{r}
#| label: data-preparation
source("scripts/01-data_preparation.R")
```

```{r}
#| label: cq-summary
#| tbl-cap: "Overview of combined Cq dataset (first 10 rows)"
head(cq_combined, 10) |> knitr::kable()
```

------------------------------------------------------------------------

## Primer Efficiency Estimation

Amplification efficiency is estimated per well from the raw RFU
amplification curves following Ramakers et al. (2003). The algorithm
fits a linear regression on log~10~-transformed RFU values across a
sliding window of 5 consecutive cycles. The window with the highest R²
among those with slope ≥ 0.15 is selected as the exponential phase, and
efficiency is computed as E = 10^slope^.

This approach is appropriate for validating primer pairs in a 2^−ΔΔCq^
study; the goal is confirming that all primers amplify within the
acceptable 90–110% efficiency range (E = 1.90–2.10).

```{r}
#| label: efficiency-estimation
source("scripts/02-efficiency_estimation.R")
```

### Per-Gene Efficiency Summary

```{r}
#| label: tbl-efficiency-gene
#| tbl-cap: "Mean primer amplification efficiency per target gene across all cultivars and wells. Efficiency flag: OK = 90–110%, Acceptable = 80–90% or 110–120%."
efficiency_per_gene |>
  dplyr::select(target, n_wells, mean_efficiency, sd_efficiency, 
                efficiency_pct, mean_r2, efficiency_flag) |>
  knitr::kable(digits = 3, col.names = c(
    "Gene", "N Wells", "Mean E", "SD E", "Efficiency (%)", "Mean R²", "Flag"
  ))
```

### Per-Gene Per-Cultivar Efficiency

```{r}
#| label: tbl-efficiency-cultivar
#| tbl-cap: "Mean primer amplification efficiency per gene per cultivar."
efficiency_per_gene_per_cultivar |>
  dplyr::select(cultivar, target, n_wells, mean_efficiency, 
                sd_efficiency, efficiency_pct, efficiency_flag) |>
  knitr::kable(digits = 3, col.names = c(
    "Cultivar", "Gene", "N Wells", "Mean E", "SD E", "Efficiency (%)", "Flag"
  ))
```

------------------------------------------------------------------------

## Fold-Change Calculation (ΔΔCq)

Relative expression is calculated using the 2^−ΔΔCq^ method (Livak &
Schmittgen, 2001):

1.  Technical replicates (two wells per biological replicate) are
    averaged to yield one mean Cq per gene × biological replicate.
2.  ΔCq = Cq(GOI) − Cq(*CmCYP*), computed per biological replicate.
3.  The calibrator ΔCq is the paired 15 DAA ΔCq value for the same
    cultivar and biological replicate (matched on `cultivar + bio_rep`).
4.  ΔΔCq = ΔCq(sample) − ΔCq(calibrator).
5.  Fold-change = 2^−ΔΔCq^.

```{r}
#| label: fold-change
source("scripts/03-fold_change.R")
```

### Per-Replicate Fold-Change (First 12 Rows)

```{r}
#| label: tbl-fc-biorep
#| tbl-cap: "Fold-change per biological replicate (2^−ΔΔCq). Values are relative to 15 DAA within each cultivar."
head(results_long, 12) |>
  knitr::kable(digits = 3)
```

### Summary: Mean ± SE Fold-Change

```{r}
#| label: tbl-fc-summary
#| tbl-cap: "Mean fold-change (±SE) per cultivar × developmental stage × gene. n = 3 biological replicates."
results_summary |>
  dplyr::select(cultivar, gene, stage, n_bio_reps, mean_fc, se_fc) |>
  knitr::kable(digits = 3, col.names = c(
    "Cultivar", "Gene", "Stage", "n", "Mean FC", "SE"
  ))
```

------------------------------------------------------------------------

## Statistical Testing

One-way ANOVA followed by Tukey HSD post-hoc pairwise comparisons is
applied per cultivar × gene combination to test for significant
differences in expression across developmental stages. Fold-change
values are log~2~-transformed prior to ANOVA to stabilize variance.
Compact letter display (CLD) notation is derived from the Tukey p-values
using `multcompView`.

```{r}
#| label: statistical-test
source("scripts/04-statistical_test.R")
```

```{r}
#| label: tbl-ddcq-tukey
#| tbl-cap: "Tukey HSD pairwise comparisons between developmental stages, per cultivar and gene. Significance: *** p < 0.001, ** p < 0.01, * p < 0.05, ns = not significant."
tukey_results |>
  dplyr::select(cultivar, gene, comparison, estimate, conf.low, conf.high, 
                adj.p.value, sig) |>
  knitr::kable(digits = 4, col.names = c(
    "Cultivar", "Gene", "Comparison", "Estimate", "95% CI Low", 
    "95% CI High", "Adj. p-value", "Sig."
  ))
```

------------------------------------------------------------------------

## Expression Figure

The final figure shows mean log~2~ fold-change per cultivar and
developmental stage for each gene, presented as grouped bar charts with
SEM error bars and CLD annotations. Stage labels (15, 30, M) are
displayed below each group of bars.

```{r}
#| label: fig-expression
#| fig-cap: "Relative expression (log2 fold-change relative to 15 DAA) of four ripening-associated genes across three developmental stages in four melon cultivars. Error bars represent SEM (n = 3). Letters above bars indicate statistically homogeneous groups based on Tukey HSD (p < 0.05)."
#| fig-width: 7
#| fig-height: 5.5
source("scripts/05-expression_plot.R")
log2_fc_plot
```

# Fruit Morphology Data Analysis

The recorded fruit morphology data include fruit weight, perimeter,
diameter, as well as quality-related trait such as flesh firmness and
soluble solid content. All the quantitative data are compared across
developmental stage to see the general pattern of melon fruit ripening
process. Comparison is done through ANOVA and Tukey's Post Hoc
statistical test under the assumption that the data are normally
distributed. The result will be presented as summary statistics table as
well as line plot for quality-related fruit morphology data (soluble
solid content and flesh firmness).

```{r}
#| message: false
#| warning: false
#| include: false
source("scripts/06-morphometrics_analysis.R")
```

## Summary Statistics Table

```{r}
#| label: morpho-summary1
#| tbl-cap: "Summary statistics of fruit morphology data. Description: fw = fruitweight, vc (vertical perimeter, hc (horizontal perimeter), vd (vertical diameter), hd (horizontal diameter), rd (rind_thickness), ft (flesh thickness), firm (flesh firmness)"

library(kableExtra)
summary_tbl |> 
  select(cultivar, stage, fw, hc, vd, hd) |> 
  kable() |> 
  kable_styling(font_size = 9)
```

```{r}
#| label: morpho-summary2
#| tbl-cap: "(Cont't) Summary statistics of fruit morphology data. Description: fw = fruitweight, vc (vertical perimeter, hc (horizontal perimeter), vd (vertical diameter), hd (horizontal diameter), rd (rind_thickness), ft (flesh thickness), firm (flesh firmness), ssc (soluble solid content"

library(kableExtra)
summary_tbl |> 
  select(cultivar, stage, rd, ft, firm, ssc) |> 
  kable() |> 
  kable_styling(font_size = 9)
```

## Statistical Test

```{r}
#| label: morpho-ANOVA
#| tbl-cap: "ANOVA statistics of fruit morphology data."
morpho_anova |> 
  select(traits, cultivar, comparison, `p adj`) |> 
  knitr::kable(digit = 6)
```

## Fruit Quality Plot

```{r}
morpho_plot
```

------------------------------------------------------------------------

# References

-   Livak KJ, Schmittgen TD (2001). Analysis of relative gene expression
    data using real-time quantitative PCR and the 2^−ΔΔCT^ method.
    *Methods* 25:402–8.
    doi:[10.1006/meth.2001.1262](https://doi.org/10.1006/meth.2001.1262)

-   Ramakers C, Ruijter JM, Deprez RH, Moorman AF (2003).
    Assumption-free analysis of quantitative real-time polymerase chain
    reaction (PCR) data. *Neuroscience Letters* 339:62–66.
    doi:[10.1016/S0304-3940(02)01423-4](https://doi.org/10.1016/S0304-3940(02)01423-4)
